\graphicspath{{6homo/asy/}}
\setcounter{section}{5}

\section{Homomorphisms \& The First Isomorphism Theorem}\label{chap:homo}

The main goal of this chapter is to discuss the link between homomorphisms, normal subgroups and factor groups. The key result is the subject of Section \ref{sec:1stiso}: the First Isomorphism Theorem.

\subsection{Kernels, Images and Normal Subgroups}\label{sec:kerimage}

In Section \ref{sec:morph} we established several important facts that are worth refreshing and summarizing here.

\begin{defn}{}{}
	The \emph{kernel} and \emph{image} of a homomorphism\footnotemark{} $\phi:G\to L$ are the sets
	\[
		\ker\phi =\bigl\{g\in G:\phi(g)=e\bigr\},\qquad 
		\image\phi=\bigl\{\phi(g):g\in G\bigr\}
	\]
\end{defn}

\footnotetext{%
	All homomorphisms in this chapter (indeed outside of Section \ref{sec:morph}) are between \emph{groups}.%
}


\begin{lemm}{}{kerimrefresh}
	Suppose $\phi:G\to L$ is a homomorphism.
	\begin{enumerate}
		\item (Exercise \ref*{sec:morph}.\ref{exs:structural2}) \ $\phi$ maps identity to identity and inverse to inverse:
		\[
			\smash{\phi(e_G)=\phi(e_L)
			\quad\text{and}\quad
			\forall g\in G, \ \bigl(\phi(g)\bigr)^{-1}=\phi(g^{-1})}
		\]
		\item (Exercise \ref*{sec:morph}.\ref{exs:kernelintro}) \ $\ker\phi$ is a subgroup of $G$ and $\image \phi$ is a subgroup of $L$.
	\end{enumerate}
\end{lemm}

Do these exercises now, even if you did them earlier!


\begin{examples}{}{}
	\exstart The function $\phi(x)=2x\negthickspace\pmod 4$ defines a homomorphism $\phi:\Z\to\Z_4$ with
	\[
		\ker\phi=\{x\in\Z:2x\equiv 0\negthickspace\negthickspace\pmod 4\}=2\Z,
		\qquad
		\image\phi=\{0,2\}
	\]
	Plainly $\ker\phi$ is a subgroup of $\Z$ and $\image\phi$ is a subgroup of $\Z_4$.
	\begin{enumerate}\setcounter{enumi}{1}
	  \item Every linear map $\rT:V\to W$ between vector spaces is a homomorphism. In linear algebra its kernel is better known as its \emph{nullspace}
		\[
			\ker\rT=\mathcal N(\rT)=\bigl\{\vv\in V:\rT(\vv)=\V0\bigr\}
		\]
		If $\rT=\rL_A:\R^n\to \R^m$ is left-multiplication by a matrix $A$, then $\image\rT$ is the \emph{column space} of $A$.
	\end{enumerate}
\end{examples}


Since the groups in these examples are abelian, all subgroups are automatically normal. In fact the kernel of \emph{every} homomorphism is a normal subgroup, though the same cannot be said for images.

\begin{lemm}{}{homosub}
	If $\phi:G\to L$ is a homomorphism, then $\ker\phi\triangleleft G$.
\end{lemm}


\begin{proof}
	The kernel is a subgroup by part 2 of Lemma \ref{lemm:kerimrefresh}, so we need only check normality. For this we appeal to the conjugation criterion (Corollary \ref{cor:normalconj}). If $g\in G$ and $k\in\ker\phi$, then
	\[
		\phi\bigl(gk g^{-1}\bigr) =\phi(g)\phi(k)\phi(g)^{-1}
	  =\phi(g)\phi(g)^{-1} =e_L
	  \implies gk g^{-1}\in\ker\phi
	\]
	Note how the first equality used the homomorphism property and part 1 of the Lemma.
\end{proof}


\goodbreak


\begin{examples}{}{}
	\exstart (Recall Exercises \ref*{sec:geomgroups}.\ref{exs:subgpmatrix1} \& \ref*{sec:morph}.\ref{exs:easyhomoexdet}) \ We've seen that $\det:\rGL_n(\R)\to\R^\times$ is a homomorphism. We therefore obtain a normal subgroup
	\[
		\ker\det=\{A\in\rGL_n(\R):\det A=1\}
		\triangleleft\rGL_n(\R)
	\]
	Otherwise said, $\rSL_n(\R)$ is a normal subgroup of $\rGL_n(\R)$.
	\begin{enumerate}\setcounter{enumi}{1}
  	\item (Example \ref*{ex:cosetexs}.\ref{ex:nonnormalcosets}, cont.) \ $\phi:\Z_2\to D_3$ defined by $\phi(x)=\mu_1^x$ is a homomorphism. Certainly $\ker\phi=\{0\}\triangleleft\Z_2$, however $\image\phi=\{e,\mu_1\}$ is not a \emph{normal} subgroup of $S_3$.
	\end{enumerate}
\end{examples}


\boldsubsubsection{Kernels, Images and Describing Homomorphisms}

Since every kernel is a normal subgroup, it makes sense to ask how its cosets may be counted and distinguished.

\begin{lemm}{}{homobij}
	Let $\phi:G\to L$ be a homomorphism. There is precisely one coset of $\ker\phi$ for each element of $\image\phi$:
	\[
		g_1\ker\phi=g_2\ker\phi
		\iff \phi(g_1)=\phi(g_2)
	\]
	Otherwise said, the \emph{index} equals the order of the image: $(G:\ker\phi)=\nm{\image\phi}$.
\end{lemm}

The proof is an abstraction of Example \ref*{ex:indexeasy}.\ref{ex:indexeasy2} (where $\phi=\det:\rO_2(\R)\to\R^\times$).

\begin{proof}
	For all $g_1,g_2\in G$, we have
	\begin{align*}
		g_1\ker\phi=g_2\ker\phi&\iff g_2^{-1}g_1\in\ker\phi \tag{Theorem \ref{thm:cosetpart}}\\
		&\iff\phi(g_2^{-1}g_1)=e_L\tag{Definition of $\ker\phi$}\\
		&\iff \phi(g_2)^{-1}\phi(g_1)=e_L\tag{Homomorphism properties}\\
		&\iff \phi(g_1)=\phi(g_2)\tag*{\qedhere}
	\end{align*}
\end{proof}

The Lemma is really part of the upcoming punchline in Section \ref{sec:1stiso}. For the moment we apply it to help describe and count homomorphisms.

\begin{thm}{}{}
	Let $\phi:G\to L$ be a homomorphism. If $G$ is finite, then $\image\phi$ is a finite group whose order divides that of $G$. The same holds for $L$. Otherwise said,
	\[
		\nm G<\infty
		\implies\nm{\image\phi}\Bigm| \nm G
		\quad\text{and}\quad 
		\nm L<\infty\implies\nm{\image\phi}\Bigm| \nm L
	\]
	If both groups are finite, then $\smash[t]{\nm{\image\phi}\Bigm|\gcd\bigl(\nm G,\nm L)}$.
\end{thm}

\begin{proof}
	If $G$ is a finite group, then $\ker\phi\le G$ is finite. Apply Lemma \ref{lemm:homobij} to see that
	\[
		\nm{\image\phi}=(G:\ker\phi) =\smash[b]{\frac{\nm G}{\nm{\ker\phi}}}
	\]
	is a divisor of $\nm G$. The second situation $\nm{\image\phi}\Bigm|\nm{L}$ is Lagrange's Theorem (\ref{thm:lagrange}).
\end{proof}


\goodbreak


\begin{examples}{}{}
	\exstart If $\phi:G\to L$ is a homomorphism and $\gcd\bigl(\nm G,\nm L\bigr)=1$, then we have $\nm{\image\phi}=1$. The image is necessarily the \emph{trivial subgroup} $\image\phi=\{e_L\}$ and there is exactly one homomorphism, namely the \emph{trivial homomorphism} ($\forall g\in G,\ \phi(g)=e_L$).\vspace{-2pt}
	\begin{enumerate}\setcounter{enumi}{1}
	  \item Consider all homomorphisms $\phi:\Z_4\to S_3$. Since the domain is cyclic, describing $\phi(1)$ is enough to obtain $\phi(x)=\bigl(\phi(1)\bigr)^x$. Since $\gcd(\nm{\Z_4},\nm{S_3})=2$, we have $\nm{\image\phi}=1$ or 2. Either:
	  \begin{itemize}
	    \item $\image\phi=\{e\}$ and we obtain the trivial homomorphism $\phi_0(x)=e$.
	    \item $\image\phi$ is a subgroup of order 2, of which $S_3$ has exactly three: $\{e,(2\,3)\}$, $\{e,(1\,3)\}$, $\{e,(1\,2)\}$. This results in three further homomorphisms (for a total of four)
	  \[
	  	\phi_1(x)=(2\,3)^x,\qquad 
	  	\phi_2(x)=(1\,3)^x,\qquad
	  	\phi_3(x)=(1\,2)^x
	  \]
	  \end{itemize}
	\end{enumerate}
\end{examples}

We now turn to the general question of homomorphisms between finite cyclic groups $\phi:\Z_m\to\Z_n$. Two facts make this relatively simple:
\begin{enumerate}\itemsep2pt
  \item As above, choosing $\phi(1)$ defines the homomorphism: $\phi(x)=\phi(1)+\cdots+\phi(1)=\phi(1)\cdot x$.
  \item $\nm{\image\phi}$ must divide $d=\gcd(m,n)$. Since $\Z_n$ has exactly one subgroup of each order dividing $n$ (Corollary \ref{cor:subscyclic}), $\image\phi$ must be a subgroup of the \emph{unique} subgroup $\ip{\frac nd}\le\Z_n$ of order $d$:
	\[
		\image\phi\le\ip{\frac nd}
		=\left\{0,\frac nd,\frac{2n}d,\ldots,\frac{(d-1)n}d\right\}
	\]
	It is enough to see what happens if we let $\phi(1)$ be each element of this subgroup in turn\ldots
\end{enumerate} 

\begin{cor}{}{homoszmn}
	There are $d=\gcd(m,n)$ distinct homomorphisms $\phi:\Z_m\to\Z_n$, namely
	\[
		\phi_k(x)=\frac{kn}{d}x
		\quad\text{where}\quad 
		k=0,\ldots,d-1
	\]
\end{cor}

\begin{proof}
	We check that each $\phi_k$ is well-defined; the calculation uses the fact that $\textcolor{blue}{\frac md}$ is an \emph{integer.} 
	\begin{align*}
		x=y\in\Z_m
		&\implies y=x+\lambda m \text{ for some $m\in\Z$}\\
		&\implies \phi_k(y)
		\begin{array}[t]{@{\;}l@{}}
			=\phi_k(x+\lambda m)
			=\dfrac{kn}d(x+\lambda m) 
			=\dfrac{kn}dx+\lambda k\textcolor{blue}{\dfrac md} n
			=\dfrac{kn}dx\\
			=\phi_k(x)
		\end{array}
		\tag{in $\Z_n$}
	\end{align*}
	By observation 1, each $\phi_k$ is a homomorphism. Observation 2 says there are no other candidates.
\end{proof}


\begin{example}{}{homoz12-20}
	We describe all homomorphisms $\phi:\Z_{12}\to\Z_{20}$.\smallbreak
	Since $\gcd(12,20)=4$, we see that $\image\phi\le\ip{5}=\{0,5,10,15\}\le\Z_{20}$. There are four choices:
	\[
		\phi_0(x)=0,\qquad 
		\phi_1(x)=5x,\qquad 
		\phi_2(x)=10x,\qquad 
		\phi_3(x)=15x\pmod{20}
	\]
	We similarly see that there are four distinct homomorphisms $\psi:\Z_{20}\to\Z_{12}$:
	\[
		\psi_0(x)=0,\qquad 
		\psi_1(x)=3x,\qquad 
		\psi_2(x)=6x,\qquad 
		\psi_3(x)=9x\pmod{12}
	\]
\end{example} 

Most of these examples rely on the domain of a homomorphism being cyclic (observation 1). It is typically much harder to find and describe homomorphisms when the domain is non-cyclic.

\goodbreak


\begin{exercises}
	Key concepts:\quad \emph{$(G:\ker\phi)=\nm{\image\phi}$\quad $\nm{\image\phi}\Bigm|\gcd\bigl(\nm G,\nm L\bigr)$}
	\begin{enumerate}
	  \item Check that you have a homomorphism (use Corollary \ref{cor:homoszmn}) and compute its kernel and image.
	  \begin{enumerate}
	    \item $\phi:\Z_{8}\to\Z_{14}$ defined by $\phi(x)=7x\pmod{14}$.
	    
	    \item $\phi:\Z_{36}\to\Z_{20}$ defined by $\phi(x)=5x\pmod{20}$.
		\end{enumerate}
	
		
		\item Describe all homomorphisms between the groups:
		\begin{enumerate}
		  \item \makebox[160pt][l]{$\phi:\Z_{15}\to\Z_{80}$\hfill (b)}\lstsp $\phi:\Z\to\Z_3$ 
		  \item[(c)] \makebox[160pt][l]{$\phi:\Z_6\to D_4$\hfill (d)}\lstsp $\phi:\Z_{15}\to A_4$
		\end{enumerate}
		
		
		%\item *Verify that $\phi:\Z\to\Z_n:x\mapsto x\pmod n$ is a homomorphism. What is its kernel? 
			
			
		\item State a linear map $\rT:\R^2\to\R^2$ whose kernel is the $y$-axis (a normal subgroup of $\R^2$). What type of linear map is the function $\rT$?
		
		
		\item Find the kernel and image of each homomorphism and verify that $\ker\phi$ is normal:
		\begin{enumerate}
		  \item The \emph{trace} of a matrix $\tr:M_2(\R)\to\R:
		  \smash{
			  \begin{smatrix}
					a&b\\c&d
			  \end{smatrix}
		  }
		  \mapsto a+d$.
			
			\item $\rT:\R^3\to\R^4:\vx\mapsto 
			\begin{smatrix}
				1&1&-1\\
				0&3&-1\\
				1&4&-2\\
				2&5&-3
			\end{smatrix}
			\vx$
		\end{enumerate}
		
		
		\item Explain why the map $\phi$ is a homomorphism and find $\ker\phi$:
		\[
			\phi:S_n\to \bigl(\{1,-1\},\cdot\bigr):\sigma \mapsto 
			\begin{cases}
	  		1&\text{ if $\sigma$ even}\\
	  		-1&\text{ if $\sigma$ odd}
			\end{cases}
		\]
		
	
		\item Suppose $\phi:G\to L$ is a homomorphism, and that $H\le G$ and $K\le L$ are subgroups.
		\begin{enumerate}
		 	\item Prove that $\phi(H):=\{\phi(h):h\in H\}$ is a subgroup of $\image\phi$.
		  \item Give an example to show that $\image\phi$ need not be a normal subgroup of $L$.
		  \item Prove that the \emph{inverse image} $\phi^{-1}(K)=\{g\in G:\phi(g)\in K\}$ is a subgroup of $G$.
		\end{enumerate}
	
		
		\item\label{exs:totient} (Exercise \ref*{sec:cyclicclass}.\ref{exs:znunits}, cont.) \ Prove that the number of distinct \emph{isomorphisms} $\phi:\Z_n\to\Z_n$ equals the order of the multiplicative group of units $(\Z_n^\times,\cdot_n)$.
% 		\[
% 			\nm{\Z_n^\times}=\nm{\bigl\{x\in\Z_n:\gcd(x,n)=1\bigr\}}
% 		\]
		
		
		\item\label{exs:homoadditive}\begin{enumerate}
		  \item Suppose $\phi:\Z_{(m)}\times\Z_{(n)}\to L$ is a well-defined homomorphism.\footnotemark{} Prove that $\phi(x,y)=ax+by$, where $a=\phi(1,0)$ and $b=\phi(0,1)$.
		  			  
	  	\item Prove that $\phi:\Z_m\times\Z_n\to\Z_m\times\Z_n$ is a well-defined homomorphism if and only if there exist integers $a,b,c,d$ for which
	  	\[
	  		\phi(x,y)=\bigl(ax+by,cx+dy\bigr),\quad 
	  		m\mid bn\quad
	  		\text{and}\quad n\mid cm
	  	\]
		\end{enumerate}
	  
	
		\item Find all homomorphisms $\phi:\Z_2\times\Z_7\to\Z_2\times\Z_5$. How do you know that there are no more?
				  
				  
 		\item Consider $\phi:D_4\to D_4:\sigma\mapsto \sigma^2$. Explain why $\phi$ is \emph{not} a homomorphism.
	\end{enumerate}
\end{exercises}


\footnotetext{%
	As in Theorem \ref{thm:cyclicisomorph}, $\Z_{(m)}$ is a generic additive cyclic group: either $\Z$ or $\Z_m$.%
}


\clearpage


\subsection{The First Isomorphism Theorem}\label{sec:1stiso}

Lemma \ref{lemm:homosub} says that every kernel is a normal subgroup. In fact \emph{all} normal subgroups arise this way.

\begin{thm}{Canonical Homomorphism}{}
	Let $G$ be a group and $H\triangleleft G$. The function
	\[
		\gamma:G\to \quotient GH
		\quad\text{defined by}\quad 
		\gamma(g)=gH
	\]
	is a homomorphism with $\ker\gamma=H$.
\end{thm}

\begin{proof}
	Since $H$ is normal, $\quotient GH$ is a factor group. By the definition of multiplication in $\quotient GH$,
	\[
		\gamma(g_1)\gamma(g_2)
		=g_1H\cdot g_2H
		=(g_1g_2)H
		=\gamma(g_1g_2)
	\]
	whence $\gamma$ is a group homomorphism. Moreover, the identity in the factor group is $H$, whence
	\[
		g\in \ker\gamma \iff \gamma(g)=H \iff g\in H 
		\tag*{\qedhere}
	\]
\end{proof}

This might feel a little sneaky, and we'd have preferred a codomain that wasn't a factor group! Our hope is vain however, for the next result---arguably the most important in elementary group theory---shows that \textbf{every homomorphism with the same kernel is essentially $\gamma$ in disguise.}

\begin{thm}{1\st{} Isomorphism}{}
	Let $\phi:G\to L$ be a homomorphism with kernel $H$. Then
	\[
		\mu:\quotient GH\to\image\phi
		\quad \text{defined by}\quad 
		\mu(gH)=\phi(g)
	\]
	is an isomorphism. Otherwise said: $\phi=\mu\circ\gamma$ and $\quotient G{\ker\phi}\cong\image\phi$.
\end{thm}


\begin{minipage}[t]{0.7\linewidth}\vspace{-5pt}
	A \emph{commutative diagram} provides a helpful summary: $\phi=\mu\circ\gamma$ means we get the same result by following arrows from $G$ to $\image\phi$ regardless of the path.\medbreak
	The 1\st{} isomorphism theorem has analogues in several other parts of mathematics: at the very least you should have met its close kin from linear algebra, the rank--nullity theorem.
\end{minipage}
\hfill
\begin{minipage}[t]{0.27\linewidth}\vspace{-5pt}
	\flushright$\xymatrix{%
		G \ar[rr]^{\phi} \ar[dr]_\gamma && \image\phi\\
		& \quotient G{\ker\phi} \ar[ur]_{\mu} &
	}$
\end{minipage}
\smallbreak




\begin{proof}
	The factor group exists since $\ker\phi\triangleleft G$ (Lemma \ref{lemm:homosub}). We check the isomorphism properties:
  \begin{description}
		\item[\normalfont\emph{Well-definition and Bijectivity}:] These are immediate after writing $H=\ker\phi$:
		\[
			g_1H=g_2H
			\underset{\text{\ref{lemm:homobij}}}{\overset{\text{Lemma}}{\iff}} \phi(g_1)=\phi(g_2)
			\iff \mu(g_1H)=\mu(g_2H)
		\]
		\item[\normalfont\emph{Homomorphism}:] For all $g_1H,g_2H\in\quotient GH$,
		\begin{align*}
			\mu(g_1H\cdot g_2H)
			&=\mu(g_1g_2H)=\phi(g_1g_2)
				\tag{coset multiplication \& definition of $\mu$}\\ 
			&=\phi(g_1)\phi(g_2)
				\tag{$\phi$ is a homomorphism}\\
			&=\mu(g_1H)\mu(g_2H)
				%\tag{definition of $\mu$ again}\\[-10pt]
				%&
				\phantom{bob}\tag*{\qedhere}
		\end{align*}
  \end{description}
	%We conclude that $\mu$ is an isomorphism.
\end{proof}


\begin{examples}{}{1stisoexs}
	\exstart Let $\phi:\Z_{12}\to\Z_{20}$ be the homomorphism $\phi(x)=5x\!\!\pmod{20}$ (Example \ref{ex:homoz12-20}). Its kernel and image are
	\begin{gather*}
		\ker\phi
		=\bigl\{x\in\Z_{12}:5x\equiv 0\negthickspace\pmod{20}\bigr\}
		=\{0,4,8\}
		=\ip 4\le \Z_{12}\\
		\image\phi=\{5x\in\Z_{20}:x\in\Z_{12}\}
		=\{0,5,10,15\}
		=\ip{5}\le\Z_{20}
	\end{gather*}
	The relevant factor group is
	\begin{align*}
		\quotient{\Z_{12}}{\ker\phi}
		&=\Bigl\{\{0,4,8\},\{1,5,9\},\{2,6,10\},\{3,7,11\}\Bigr\}
		=\bigl\{\ip 4,1+\ip 4,2+\ip 4,3+\ip 4\bigr\}
	\end{align*}
	The canonical homomorphism $\gamma$ and the isomorphism $\mu$ are\vspace{-8pt}
	\begin{enumerate}\setcounter{enumi}{1}
	  \item[]\begin{quote}
			$\SelectTips{cm}{}\xymatrix @C40pt @R0pt{%
				\rlap{$\gamma(x)=x+\ip 4$} &&& \Z_{12} \ar@{->}[r]_-{\scriptstyle\gamma} \ar@{->}@/^12pt/[rr]^{\scriptstyle\phi} & \quotient{\Z_{12}}{\ip 4} \ar@{->}[r]_-{\scriptstyle\mu} & \image\phi\\
				\rlap{$\mu\bigl(x+\ip 4\bigr)=5x$} &&& x \ar@{|->}[r] & x+\ip 4 \ar@{|->}[r] & 5x
			}$
		\end{quote}
		

	  
	\item The homomorphism $\phi:\R\to(\C^\times,\cdot):x\mapsto e^{2\pi ix}$ has
  \[
  	\ker\phi=\bigl\{x\in\R:e^{2\pi ix}=1\bigr\} =\Z
  	\quad\text{and}\quad 
  	\image\phi=S^1
  	\tag{$S^1$ is the circle group}
  \]
  The canonical homomorphism $\gamma$ and the isomorphism $\mu$ from the theorem are
  \[
  	\gamma:\R\to\quotient{\R}{\Z}:x\mapsto x+\Z
  	\quad\text{and}\quad
  	\mu:\quotient\R{\Z}\to S^1:x+\Z\mapsto e^{2\pi ix}
  \]
  Think about how this relates to Example \ref*{ex:indexeasy}.\ref{ex:ZsubgroupRindex}. 
  
  \item\label{ex:1stisoexs3} The homomorphism $\phi:\Z\times\Z\to\Z:(x,y)\mapsto 3x-2y$ satisfies
  \[
  	\phi(x,y)=(0,0)\iff 3x=2y\iff (x,y)=(2n,3n)\text{ for some $n\in\Z$}
  \]
  We conclude that $\ker\phi=\ip{(2,3)}$. The canonical homomorphism is therefore
  \[
  	\gamma:\Z\times\Z\to
  	\quotient{\Z\times\Z}{\ip{(2,3)}}:
  	(x,y)\mapsto (x,y)+\ip{(2,3)}
  \]
  Since $n=\phi(n,n)$, the homomorphism is surjective: $\image\phi=\Z$. By the 1\st{} isomorphism theorem,
  \[
  	\quotient{\Z\times\Z}{\ip{(2,3)}}\cong\Z
  	\quad\text{via}\quad 
  	\mu\bigl((x,y)+\ip{(2,3)}\bigr)=3x-2y
  \]
  
  \item $\phi(x,y)=(x,y-x)$ is a well-defined homomorphism $\phi:\Z\times\Z_6\to\Z_3\times\Z_6$. Moreover,
  \begin{gather*}
  	\phi(x,y)=(0,0)\iff
  	\begin{cases}
  		x=3k,\text{ and}\\
  		y=x=3k\in\Z_6
  	\end{cases}\\
  	\text{whence}\quad
  	\ker\phi=\ip{(3,3)}=\bigl\{\ldots,(-6,0),(-3,3),(0,0),(3,3),(6,0),\ldots\bigr\}
  \end{gather*}
  Moreover, $\phi$ is surjective: e.g. $(m,n)=\phi(m,n+m)$. We conclude that
	\[
		\quotient{\Z\times\Z_6}{\ip{(3,3)}}\cong\Z_3\times\Z_6
		\quad\text{via}\quad
		\mu\bigl((x,y)+\ip{(3,3)})=(x,y-x)
	\]
	\end{enumerate}
\end{examples}

\vfil

\goodbreak

\boldsubsubsection{Identifying Factor Groups (revisited)}

The 1\st{} isomorphism theorem may be applied to the identification of factor groups. Given $H\triangleleft G$, we cook up a homomorphism $\phi:G\to L$ with $\ker\phi=H$ and conclude that $\quotient GH\cong\image\phi$. This typically requires some creativity: there are many options, and both $L$ and a formula for $\phi$ need to be found simultaneously! We first revisit some examples from the previous section in this context.

\begin{examples*}{\ref{ex:factorfiniteabelian},\,mk.II}{}
	For each subgroup $H$ of $G=\Z_4\times\Z_8$, we hunt for a suitable homomorphism with $\ker\phi =H$.
	\begin{enumerate}
	  \item Given $H=\ip{(0,1)}$, we need a homomorphism for which $\phi(0,1)$ is the identity. This is very simple: just ignore $y$! Define
	  \[
	  	\phi:\Z_4\times\Z_8\to\Z_4:(x,y)\mapsto x
	  \]
	  This indeed has kernel $\ker\phi=\{(0,y):y\in\Z_8\} =H$. By the 1\st{} isomorphism theorem,
	  \[
	  	\quotient GH\cong\image\phi=\Z_4
	  	%\quad\text{via}\quad
	  	%\mu:(x,y)+H\mapsto x
	  \]
	  via the isomorphism $\mu\bigl((x,y)+H\bigr)=x$. Note that $(x,y)+H=(x,0)+H$, whence $\mu$ is the \emph{inverse} of the isomorphism $\psi:x\mapsto (x,0)+H$ stated previously!
	
	  \item Given $H=\ip{(0,2)}$ we require $\phi(0,2)$ to be the identity. This may be achieved by taking $y$ modulo 2 and defining
	  \[
	  	\phi:\Z_4\times\Z_8\to\Z_4\times\Z_2:(x,y)\mapsto (x,y)
	  \]
	  This is well-defined
	  \[
	   	\phi(x+4m,y+8n)=(x+4m,y+8n)=(x,y)=\phi(x,y)\tag{since $2\mid 8$}
	  \]
	  and a homomorphism with required kernel $H$. Moreover, $\phi$ is surjective, whence
	  \[
	  	\quotient GH\cong \image\phi= \Z_4\times\Z_2
	  \]
	  via the isomorphism $\mu\bigl((x,y)+H\bigr) =(x,y)$. Once again $\mu$ is the inverse of $\psi(x,y)=(x,y)+H$ in the original example.
	  
	  \item Finding a homomorphism with kernel $H=\ip{(2,4)}=\bigl\{(0,0),(2,4)\bigr\}$ is somewhat trickier. One approach is to observe that
	  \[
	  	(x,y)\in H\iff x\equiv 0\negthickspace\negthickspace\pmod 2
	  	\ \text{ and }\ 
	  	y-2x\equiv 0\negthickspace\negthickspace\pmod 8
	  \]
	  This suggests
	  \[
	 		\phi:\Z_4\times\Z_8\to\Z_2\times\Z_8:(x,y)\mapsto\bigl(x,y-2x\bigr)
	 	\]
	  It is worth checking that this is \textcolor{red}{well-defined}: the $\textcolor{red}{2}x$ in the second co-ordinate is crucial! Certainly $\phi$ has the correct kernel. It is moreover surjective: $(m,n)=\phi(m,n+2m)$. In conclusion,
	  \[
	  	\quotient GH\cong\image\phi=\Z_2\times\Z_8
	  \]
	  via the isomorphism $\mu\bigl((x,y)+H\bigr)=(x,y-2x)$.
	\end{enumerate}
\end{examples*}


\begin{examples}{}{1stisonasty}
	Two final examples follow a similar theme, though with infinite groups. We'll generalize these in Exercise \ref{exs:genZZfactor}.
	\begin{enumerate}
	  \item We identify the factor group $\quotient GH=\quotient{\Z\times\Z}{\ip{(1,-1)}}$.\smallbreak
	  We require a homomorphism where $(1,-1)\in\ker\phi$. An obvious candidate is
	  \[
	  	\phi:\Z\times\Z\to\Z:(x,y)\mapsto x+y
	  \]
	  Certainly $(x,y)\in\ker\phi\Longleftrightarrow x+y=0\iff y=-x\iff (x,y)\in\ip{(1,-1)}$. Moreover, $x=\phi(x,0)$ for any $x\in\Z$, so this is a surjective homomorphism. We conclude that
	  \[
	  	\quotient{\Z\times\Z}{\ip{(1,-1)}}\cong\image\phi=\Z
	  \]
	  
	  \item We identify the factor group $\quotient GH=\quotient{\Z\times\Z}{\ip{(4,6)}}$.\smallbreak
	  It is tempting to try a homomorphism $\psi(x,y)=6x-4y$ or $\psi(x,y)=3x-2y$. These certainly satisfy $\psi(4,6)=0$, but they also have $\psi(2,3)=0$ which we don't want!\smallbreak
	  As a hint for how to proceed, note that the element/coset $(2,3)+H\in\quotient GH$ has order 2:
	  \[
	  	2(2,3)=(4,6)\in H
	  \]
	  We'd therefore like a homomorphism that maps $(2,3)$ to an element of order 2: the simple function $\xi:\Z\times\Z\to\Z_2:(x,y)\mapsto y-x$ certainly does this! We try combining these functions:
	  \[
	  	\phi:\Z\times\Z\to \Z\times\Z_2:(x,y)\mapsto (3x-2y,y-x)
	  \]
	  Let's compute the kernel:
	  \[
	  	\phi(3x-2y,y-x)=(0,0)\implies 3x=2y\in\Z \implies (x,y)=(2m,3m): m\in\Z
	  \]
	  Substitute this into the second factor: $y-x=m=0\in\Z_2\Longrightarrow m=2n$ is even, from which  $(x,y)=(4n,6n)$. It is trivially verified that all such pairs lie in the kernel, whence $\ker\phi=\ip{(4,6)}$, as required. Finally,
	  \[
	  	\phi(1,1)=(1,0),\quad \phi(2,3)=(0,1)\implies (u,v)=\phi(u+2v,u+3v)
	  \]
	  says that $\phi$ is surjective. We conclude that $\quotient GH\cong\Z\times\Z_2$.
	\end{enumerate}
\end{examples}

%\goodbreak

\begin{exercises}{}{}
	Key concepts:
	\begin{quote}
		\emph{Canonical homomorphism $\gamma:G\to\quotient GH$\qquad 1\st{} isomorphism theorem $\mu:\quotient G{\ker\phi}\cong\image\phi$}
	\end{quote}
	
	\begin{enumerate}
	 % \item Prove that $\quotient{\Q}{\Z}\cong S^1_{\Q}=\{e^{2\pi iq}:q\in\Q\}$.
	  
	  \item Let $\phi:\Z_{18}\to\Z_{12}$ be the homomorphism $\phi(x)=10x$. 
	  \begin{enumerate}
	    \item Find the kernel of and image of $\phi$.
	    \item List the elements of the factor group $\smash[b]{\quotient{\Z_{18}}{\ker\phi}}$.
	    \item State an explicit isomorphism $\mu:\quotient{\Z_{18}}{\ker\phi}\to\image\phi$.
	    \item To what basic group $\Z_n$ is the factor group isomorphic?
	  \end{enumerate}
	  
	 	
	 	\item Repeat the previous question for the homomorphism $\phi:\Z\to\Z_{20}:x\mapsto 8x$.
	  
	  
	  \item For each function $\phi:\Z\times\Z\to\Z$, find the kernel and identify the factor group $\smash{\quotient{\Z\times\Z}{\ker\phi}}$.
	  \begin{enumerate}
	    \item $\phi(x,y)=3x+y$\qquad\qquad 
	    (b)\lstsp $\phi(x,y)=2x-4y$
	  \end{enumerate}  
	  
	    
	  \item\begin{enumerate}
	  	\item If a subgroup $H$ of $G=\Z_{15}\times\Z_3$ has order 5, find its elements.
			\item Show that $\phi(x,y)=(x,y)$ is a homomorphism $\phi:G\to\Z_3\times\Z_3$ with $\ker\phi=H$.
			\item What does the 1\st{} isomorphism theorem tell us about the factor group $\quotient GH$?
		\end{enumerate}
	  
	  
	  \item Suppose $G$ is a finite group with normal subgroup $H$ and that $\phi:G\to L$ is a homomorphism with $\ker\phi=H$. Prove that $(G:H)\le \nm L$ with equality if and only if $\phi$ is surjective.
	  
	  
	  \item Consider the map $\phi:\Z\times\Z_{12}\to \Z_3\times\Z_6$ defined by
	  \[
	  	\phi(x,y)=\bigl(2x+y,y\bigr)
	  \]
	  \begin{enumerate}
	    \item Verify that $\phi$ is a well-defined homomorphism.
	    \item Compute $\ker\phi$ and identify the factor group $\quotient{\Z\times\Z_{12}}{\ker\phi}$
	  \end{enumerate}
	  
	  
	  \item Let $H=\ip{(3,1)}\le G=\Z_9\times\Z_3$. Find an explicit homomorphism $\phi:G\to\Z_9$ whose kernel is $H$, and thus identify the factor group $\quotient GH$.\par
	  (\emph{Hint: $(x,y)\in H=\{(0,0),(3,1),(6,2)\}\iff\ldots$})
	  
	  \item Consider $H=\ip{(3,3)}\le G=\Z_9\times\Z_9$. Find a surjective homomorphism $\phi:G\to\Z_3\times\Z_9$ whose kernel is $H$ and hence prove that $\quotient GH\cong\Z_3\times\Z_9$.
	  
	  
	  \item Let $\phi:S^1\to S^1:z\mapsto z^2$.
	  \begin{enumerate}
	    \item Find the kernel of $\phi$ and describe the canonical homomorphism $\gamma:S_1\to\smash{\quotient{S^1}{\ker\phi}}$.
	    \item What does the first isomorphism theorem say about the factor group $\smash{\quotient{S^1}{\ker\phi}}$.
	    \item For each $n$, identify the factor group $\smash{\quotient{S^1}{U_n}}$, where $U_n$ is the group of $n\th$ roots of unity.
	  \end{enumerate}
	 
	  %\item Recall Example \ref*{ex:normalmoreex}.\ref{ex:nastyiso}. If $H=\ip{(1,2,2)}\le G=\Z_5\times\Z_6\times\Z$, try to find a surjective homomorphism $\phi:G\to \Z_5\times\Z_6\times\Z_2$ with $\ker\phi=H$.
	  
	  
	  \item\label{exs:genZZfactor} We identify $\quotient GH=\quotient{\Z\times\Z}{\ip{(a,b)}}$. Suppose $d=\gcd(a,b)$ where $a,b$ are not both zero.
	  \begin{enumerate}
	    \item Let $\psi(x,y)=\frac bd x-\frac ad y$ be the homomorphism $\psi:\Z\times\Z\to\Z$. Find $\ker\psi$.
	    \item Show that $\quotient GH$ contains an element of order $d$, namely $(\frac ad,\frac bd)+H$.
	    \item Bézout's identity says $\exists\kappa,\lambda\in\Z$ for which $\kappa a+\lambda b=d$. The function
	    \[
	    	\xi:\Z\times\Z\to\Z_d:(x,y)\mapsto \kappa x+\lambda y
	    \]
	    plainly maps $(\frac ad,\frac bd)$ to the generator $1\in\Z_d$. Compute $\ker\phi$ for the combined map
			\[
				\phi:\Z\times\Z\to\Z\times\Z_d:(x,y)\mapsto \left(\frac bd x-\frac ad y,\kappa x+\lambda y\right)
			\]
			\item Identify the factor group $\quotient{\Z\times\Z}{\ip{(a,b)}}$.
			\item Find an explicit isomorphism $\mu:\quotient{\Z\times\Z}{\ip{(21,35)}}\to\Z\times\Z_7$.
		\end{enumerate}
	\end{enumerate}
\end{exercises}




\clearpage


\subsection{Conjugation, Cycle Types, Centers and Automorphisms}\label{sec:conj}

In this section we consider an important type of homomorphism and some its consequences.

\begin{defn}{}{}
	Let $G$ be a group and $x,y\in G$. We say that $y$ is \emph{conjugate to} $x$ if
	\[
		\exists g\in G\quad\text{such that}\quad y=gx g^{-1}
	\]
	If $g\in G$ is fixed, then \emph{conjugation by $g$} is the map $c_g:G\to G:x\mapsto gx g^{-1}$.
\end{defn}

We've met this notion before: recall that a subgroup $H$ is normal if and only if $c_g(h)\in H$ for all $g\in G$ (Corollary \ref{cor:normalconj}). It should also be familiar from linear algebra, in the context of \emph{similarity.} Recall that square matrices $A,B$ are similar if $B=MAM^{-1}$ for some invertible $M$. Such matrices have the same eigenvalues and, essentially, `do the same thing' with respect to different bases. An explicit group theory analogue of this is Theorem \ref{thm:cycletype} below.


\begin{lemm}{}{conjiso}
	Conjugation by $g$ is a isomorphism $c_g:G\cong G$.
\end{lemm}

\begin{proof}
	Conjugation by $g^{-1}$ is the inverse function of $c_g^{-1}$:
	\[
		c_{g^{-1}}\bigl(c_g(x)\bigr)=g^{-1}gx g^{-1}(g^{-1})^{-1}=x,\ \text{ etc.}
	\]
	We moreover have a homomorphism:
	\[
		c_g(xy)=g(xy)g^{-1}
		=\bigl(gx g^{-1}\bigr)\bigl(gy g^{-1}\bigr)
		=c_g(x)c_g(y)
		\tag*{\qedhere}
	\]
\end{proof}


\begin{lemm}{}{conjequiv}
	Conjugacy is an equivalence relation \ ($x\sim y\iff \exists g\in G$ such that $y=gx g^{-1}$).
\end{lemm}

The proof is an exercise. The equivalence classes under conjugacy are termed \emph{conjugacy classes.}


\begin{examples}{}{}
	\exstart If $G$ is abelian then every conjugacy class contains only one element:
	\[
		x\sim y\iff \exists g\in G
		\quad\text{such that}\quad 
		y=gx g^{-1}=xg g^{-1}=x
	\]
	\begin{enumerate}\setcounter{enumi}{1}
		\item The smallest non-abelian group is $S_3$ has conjugacy classes
		\[
			\{e\},\quad \{(1\,2),(1\,3),(2\,3)\},\quad \{(1\,2\,3),(1\,3\,2)\}
		\]
		This can be computed directly, but it follows immediately from\ldots
	\end{enumerate}
\end{examples}


\begin{thm}{}{cycletype}
	The conjugacy classes of $S_n$ are the \emph{cycle types}: elements are conjugate if and only if they have the same cycle-type.
\end{thm}

If an element $\sigma\in S_n$ is written as a product of disjoint cycles, then its cycle-type is clear. For instance:
\begin{itemize}
  \item $(1\,2\,3)(4\,5)$ has the same cycle-type as $(1\,5\,6)(2\,3)$: we might call these 3,2-cycles.
  \item $(1\,2)(3\,4)$ has cycle-type 2,2.
\end{itemize}


\goodbreak


Before seeing a proof it is beneficial to try an example.

\begin{example}{}{}
If $\rho=(2\,4\,3)$ and $\sigma=(1\,2)(3\,4)$ in $S_4$, then
  \[
  	\rho\sigma\rho^{-1} =(2\,4\,3)(1\,2)(3\,4)(2\,3\,4) =(1\,4)(2\,3)
  \]
  This plainly has the same cycle-type as $\sigma$. Moreover, it may be obtained by applying $\rho$ to the entries of $\sigma$!
  \[
  	\rho\sigma\rho^{-1}=(1\,4)(2\,3) 
  	=\bigl(\rho(1)\,\rho(2)\bigr)\bigl(\rho(3)\,\rho(4)\bigr) \tag{$\ast$}
  \]
  \begin{minipage}[t]{0.75\linewidth}\vspace{-5pt}
	  This tells us how to reverse the process: given 2,2-cycles $\sigma=(1\,2)(3\,4)$ and $\tau=(1\,4)(2\,3)$ simply place $\sigma$ on top of $\tau$ in a table to define a suitable $\rho=(2\,4\,3)$ for which $\rho\sigma\rho^{-1}=\tau$.
	  \end{minipage}
	  \hfill
	  \begin{minipage}[t]{0.2\linewidth}\vspace{-5pt}
	    $\begin{array}{l||cc|cc}
	  		x&1&2&3&4\\\hline
	  		\rho(x)&1&4&2&3
	  	\end{array}$
  \end{minipage}
\end{example}


The proof is a notational horror, but it amounts to nothing more than the example done abstractly: our goal is to recognize $\rho\sigma\rho^{-1}$ in exactly the form $(\ast)$, taking $\rho$ of each entry in each cycle.

\begin{proof}
	\begin{description}
		\item[\normalfont($\Rightarrow$)] First consider conjugation of a $k$-cycle $\sigma=(a_1\cdots a_k)$ by $\rho\in S_n$. We claim that
		\[
			\rho\sigma\rho^{-1}=\bigl(\rho(a_1)\cdots\rho(a_k)\bigr)
		\]
		is \textbf{also a $k$-cycle.} Since this is an equality of functions, we verify by evaluating on all elements of $X=\{1,2,\ldots,n\}$. There are two cases, which correspond to partitioning $X=A\cup(X\setminus A)$ where $A=\{a_1,\ldots,a_k\}$. First observe that $\rho$ bijectively maps $A\to A$ and $X\setminus A\to X\setminus A$.
		\begin{enumerate}
			\item Suppose $x\in A$. Since $A=\rho(A)$, we may uniquely write $x=\rho(a_j)$. But then
			\[
				\rho\sigma\rho^{-1}(x) =\rho\sigma\rho^{-1}\bigl(\rho(a_j)\bigr)
				=\rho\sigma(a_j)=\rho(a_{j+1})
			\]
			where $a_{k+1}$ is understood to equal $a_1$.
			\item Suppose $x\not\in A$. We know that $\rho^{-1}(x)\not\in A$ is unmoved by $\sigma$, whence
			\[
				\rho\sigma\rho^{-1}(x) 
				=\rho\sigma(\rho^{-1}(x))
				=\rho\rho^{-1}(x)=x
			\]
		\end{enumerate}
		We conclude that $\rho\sigma\rho^{-1}=\bigl(\rho(a_1)\cdots\rho(a_k)\bigr)$	is as claimed.
		More generally, if $\sigma=\sigma_1\cdots\sigma_l$ is a product of disjoint cycles, then
		\[
			\rho\sigma\rho^{-1} =(\rho\sigma_1\rho^{-1})(\rho\sigma_2\rho^{-1})\cdots (\rho\sigma_l\rho^{-1})
		\]
		has the same cycle-type as $\sigma$.
		\item[\normalfont($\Leftarrow$)] Suppose $\sigma=\sigma_1\cdots\sigma_l$ and $\tau=\tau_1\cdots\tau_l\in S_n$ have the same cycle-type, written so that corresponding orbits have the same length. Assume we've all single-element orbits are included so that $\bigcup\sigma_i=\{1,\ldots,n\}=\bigcup\tau_i$. Define $\rho\in S_n$ by writing the orbits of $\sigma$ and $\tau$ on top each other
		\[
			\begin{array}{l||c|c|c|c}
	      x&\sigma_1 &\sigma_2 & \cdots & \sigma_l\\\hline
	      \rho(x)&\tau_1 & \tau_2 & \cdots & \tau_l
	    \end{array}
	  \]
		Since the orbits of $\sigma$, \& $\tau$ partition $X=\{1,\ldots,n\}$, this is plainly defines a permutation. Finally, if $s_{i,j}$ and $t_{i,j}$ are the $j\th$ elements of the orbits $\sigma_i$ and $\tau_i$, then
		\[
			\rho\sigma\rho^{-1}(t_{i,j})
			=\rho\sigma(s_{i,j})=\rho\bigl(s_{i,j+1}\bigr)
			=t_{i,j+1}=\tau\bigl(t_{i,j}\bigr)
		\]
		We conclude that $\rho\sigma\rho^{-1}=\tau$, as required.\qedhere
	\end{description}
\end{proof}


% \begin{proof}
% 	\begin{description}
% 		\item[\normalfont($\Rightarrow$)] Consider conjugation by $\rho\in S_n$. First let $\sigma=(a_1\cdots a_k)$ be a $k$-cycle and write
% 		\[
% 			A=\{a_1,\ldots,a_k\},\qquad R=\rho(A)=\bigl\{\rho(a_1),\ldots,\rho(a_k)\bigr\}
% 		\]
% 		Since $\rho$ is a bijection, $\nm R=k$ and $x\in R\Longleftrightarrow \rho^{-1}(x)\in A$. To evaluate $\rho\sigma\rho^{-1}$ as a function, we must consider two sub-cases:
% 		\begin{enumerate}
% 			\item Suppose $x\in R$. Write $x=\rho(a_j)$. But then
% 			\[
% 				\rho\sigma\rho^{-1}(x) =\rho\sigma\rho^{-1}\bigl(\rho(a_j)\bigr)
% 				=\rho\sigma(a_j)=\rho(a_{j+1})
% 			\]
% 			where $a_{k+1}$ is understood to equal $a_1$.
% 			\item Suppose $x\not\in R$. Since $\rho^{-1}(x)\not\in A$, it is unmoved by $\sigma$, whence
% 			\[
% 				\rho\sigma\rho^{-1}(x) 
% 				=\rho\sigma(\rho^{-1}(x))
% 				=\rho\rho^{-1}(x)=x
% 			\]
% 		\end{enumerate}
% 		We conclude that $\rho\sigma\rho^{-1}=\bigl(\rho(a_1)\cdots\rho(a_k)\bigr)$	is \textbf{also a $k$-cycle}!\par
% 		More generally, if $\sigma=\sigma_1\cdots\sigma_l$ is a product of disjoint cycles, then
% 		\[
% 			\rho\sigma\rho^{-1} =(\rho\sigma_1\rho^{-1})(\rho\sigma_2\rho^{-1})\cdots (\rho\sigma_l\rho^{-1})
% 		\]
% 		has the same cycle-type as $\sigma$.
% 		\item[\normalfont($\Leftarrow$)] Suppose $\sigma=\sigma_1\cdots\sigma_l$ and $\tau=\tau_1\cdots\tau_l\in S_n$ have the same cycle type, written so that corresponding orbits have the same length. Moreover, assume we've included all necessary single-element orbits so that $\bigcup\sigma_i=\{1,\ldots,n\}=\bigcup\tau_i$. Define a permutation $\rho$ by writing the orbits of $\sigma$ and $\tau$ on top each other
% 		\[
% 			\begin{array}{l||c|c|c|c}
% 	      x&\sigma_1 &\sigma_2 & \cdots & \sigma_l\\\hline
% 	      \rho(x)&\tau_1 & \tau_2 & \cdots & \tau_l
% 	    \end{array}
% 	  \]
% 		Since the orbits of $\tau$ partition $X=\{1,\ldots,n\}$, this is plainly a permutation. Finally, if $s_{i,j}$ and $t_{i,j}$ are the $j\th$ elements of the orbits $\sigma_i$ and $\tau_i$, then
% 		\[
% 			\rho\sigma\rho^{-1}(t_{i,j})
% 			=\rho\sigma(s_{i,j})=\rho\bigl(s_{i,j+1}\bigr)
% 			=t_{i,j+1}=\tau\bigl(t_{i,j}\bigr)
% 		\]
% 		We conclude that $\rho\sigma\rho^{-1}=\tau$, as required.\qedhere
% 	\end{description}
% \end{proof}


\goodbreak


\begin{examples}{}{conjexample}
	\exstart The permutations $\sigma=(1\,4\,5)(2\,7\,6)$ and $\tau=(1\,6\,5)(3\,4\,7)$ in $S_7$ have the same cycle-type and are thus conjugate. The table defines a suitable $\rho$.
	\[
		\begin{array}{l||ccc|ccc|c}
      x&1&4&5&2&7&6&3\\\hline
      \rho(x)&1&6&5&3&4&7&2
    \end{array}
    \implies \rho=(2\,3)(4\,6\,7)
   \]
	It is worth making a sanity check:
	\[
		\rho\sigma\rho^{-1}
		=(2\,3)(4\,6\,7)(1\,4\,5)(2\,7\,6)(2\,3)(4\,7\,6)
		=(1\,6\,5)(3\,4\,7) =\tau
	\]
	Other choices of $\rho$ are available: just write the orbits of $\sigma,\tau$ in different orders. Can you figure out \emph{how many} distinct choice of $\rho$ will work?
	\begin{enumerate}\setcounter{enumi}{1}
		\item\label{ex:conjexample2} (Example \ref*{sec:factor}.\ref{ex:a4normal}) \ We checked previously that $V=\bigl\{e,(1\,2)(3\,4),(1\,3)(2\,4),(1\,4)(2\,3)\bigr\}$ is a normal subgroup of $A_4$. Since $V$ contains the identity and all 2,2-cycles it is closed under conjugacy and thus a normal subgroup of both $A_4$ and $S_4$.
	\end{enumerate}
\end{examples}



\boldsubsubsection{Automorphisms}

We've already seen that conjugation $c_g:G\to G$ by a fixed element $g\in G$ is an isomorphism. We now consider all such maps.

\begin{defn}{}{}
	An \emph{automorphism} of a group $G$ is an isomorphism of $G$ with itself. The set of such is denoted $\Aut G$. The \emph{inner automorphisms} are the conjugations
	\[
		\Inn G=\{c_g:G\to G\text{ where }c_g(x)=gx g^{-1}\}
	\]
\end{defn}

\begin{example}{}{}
	There are four homomorphisms $\phi_k:\Z_4\to\Z_4$ (Corollary \ref{cor:homoszmn});
	\[
		\phi_0(x)=0,\quad 
		\phi_1(x)=x,\quad 
		\phi_2(x)=2x,\quad 
		\phi_3(x)=3x
	\]
	of which two are automorphisms: $\Aut\Z_4=\{\phi_1,\phi_3\}$.
	Observe that $\phi_1$ is the identity function and that $\phi_3\circ\phi_3=\phi_1$. The automorphisms therefore comprise a \emph{group} (isomorphic to $\Z_2$) under composition of functions.\smallbreak
	As for conjugations, since $\Z_4$ is abelian these are uninteresting:
	\[
		g,x\in\Z_4\implies 
		c_g(x)=g+x+(-g)=x
		\implies c_g=\phi_1
	\]
	There is only one inner automorphism of $\Z_4$, namely the identity function $\phi_1$. This indeed holds in general: if $G$ is abelian, then $\Inn(G)\cong\Z_1$ is trivial.
\end{example}

In general, hunting for automorphisms is difficult. Here is a simple observation that helps to narrow things down. The proof is an exercise.

\begin{lemm}{}{autoorder}
	If $\phi\in\Aut G$ and $x\in G$, then the orders of $x$ and $\phi(x)$ are identical.
\end{lemm}

This helps to streamline the previous example: $\phi(1)$ must have the same order (four) as 1 and so our only possibilities are $\phi(1)=1$ or $\phi(1)=3$. These generate the two observed automorphisms.


\goodbreak


\begin{example}{}{}
	We describe all automorphisms $\phi$ of $S_3$. Consider $\sigma=(1\,2)$ and $\tau=(1\,2\,3)$. Since the order of an element is preserved by $\phi$, we conclude that
  \[
  	\phi(e)=e,\quad 
  	\textcolor{red}{\phi(\sigma)}\in\bigl\{(1\,2),(1\,3),(2\,3)\bigr\},\quad 
  	\textcolor{blue}{\phi(\tau)}\in \bigl\{(1\,2\,3),(1\,3\,2)\bigr\}
 	\]
  We therefore have a maximum of \emph{six} possible automorphisms; it is tedious to check, but \emph{all} in fact define automorphisms! Indeed all are conjugations, from which $\Aut S_3=\Inn S_3$. Here is the data; verify some of it for yourself:\vspace{-2pt}
  \[
	  \begin{array}{c||cccccc}
		  \text{element $g$}&c_g(e)&\textcolor{red}{c_g(1\,2)}&c_g(1\,3)&c_g(2\,3)&\textcolor{blue}{c_g(1\,2\,3)}&c_g(1\,3\,2)\\\hline\hline
		  e&e&\textcolor{red}{(1\,2)}&(1\,3)&(2\,3)&\textcolor{blue}{(1\,2\,3)}&(1\,3\,2)\\\hline
		  (1\,2)&e&\textcolor{red}{(1\,2)}&(2\,3)&(1\,3)&\textcolor{blue}{(1\,3\,2)}&(1\,2\,3)\\\hline
		  (1\,3)&e&\textcolor{red}{(2\,3)}&(1\,3)&(1\,2)&\textcolor{blue}{(1\,3\,2)}&(1\,2\,3)\\\hline
		  (2\,3)&e&\textcolor{red}{(1\,3)}&(1\,2)&(2\,3)&\textcolor{blue}{(1\,3\,2)}&(1\,2\,3)\\\hline
		  (1\,2\,3)&e&\textcolor{red}{(2\,3)}&(1\,2)&(1\,3)&\textcolor{blue}{(1\,2\,3)}&(1\,3\,2)\\\hline
		  (1\,3\,2)&e&\textcolor{red}{(1\,3)}&(2\,3)&(1\,2)&\textcolor{blue}{(1\,2\,3)}&(1\,3\,2)
	  \end{array}
  \]
  As the next result shows, the automorphisms always form a group under composition; in this case $\Aut S_3$ is a group of order 6 which is easily seen to be \emph{non-abelian}, 
  \[
  	c_{(1\,2)}c_{(1\,3)}= c_{(1\,3\,2)}\neq c_{(1\,2\,3)}= c_{(1\,3)}c_{(1\,2)}
  \]
  By process of elimination, we conclude that $\Aut S_3\cong S_3$.
\end{example}


\begin{thm}{}{aut}
	$\Aut G$ and $\Inn G$ are groups under composition. Moreover $\Inn G\triangleleft\Aut G$.
\end{thm}

\begin{proof}
	That $\Aut G$ is a group is simply the fact that composition and inverses of isomorphisms are isomorphisms: you should already have made this argument when answering Exercise \ref*{sec:morph}.\ref{exs:isomorphiccomposition}. By Lemma \ref{lemm:conjiso}, every conjugation is an isomorphism, and it is simple to check that $c_g\circ c_h=c_{gh}$ and $c_g^{-1}=c_{g^{-1}}$. We conclude that $\Inn G\le \Aut G$.\smallbreak
	For normality, we check that $\Inn G$ is itself closed under conjugation! Let $\tau\in\Aut G$ and $c_g\in\Inn G$. For any $x\in G$, we have\footnotemark
	\begin{align*}
		(\tau c_g\tau^{\textcolor{red}{-1}})(x)&=\tau\Bigl(c_g\bigl(\tau^{\textcolor{red}{-1}}(x)\bigr)\Bigr)\\
		&=\tau\Bigl(g\bigl(\tau^{\textcolor{red}{-1}}(x)\bigr)g^{\textcolor{blue}{-1}}\Bigr) \tag{definition of $c_g$}\\
		&=\Bigl(\tau(g)\Bigr)\Bigl(\tau\bigl(\tau^{\textcolor{red}{-1}}(x)\bigr)\Bigr)\Bigl(\tau(g^{\textcolor{blue}{-1}})\Bigr)\tag*{(since $\tau$ is a homomorphism)}\\
		&=\bigl(\tau(g)\bigr)x\bigl(\tau(g)\bigr)^{\textcolor{blue}{-1}}\tag*{(again since $\tau$ is an homomorphism)}\\
		&=c_{\tau(g)}(x)
	\end{align*}
	We conclude that $\tau c_g\tau^{-1}=c_{\tau(g)}\in \Inn G$, from which $\Inn G\triangleleft\Aut G$.
\end{proof}

\vspace{-2pt}

\footnotetext{%
	The challenge in reading the proof is keeping track of where everything lives. To help, the inverse symbol is colored: $\tau^{\textcolor{red}{-1}}$ (in red) means the inverse \emph{function,} whereas $g^{\textcolor{blue}{-1}}$ (in blue) means the inverse of an \emph{element} in $G$.%
}


\goodbreak


\boldsubsubsection{Centers}

We say that an element $g\in G$ \emph{commutes} with another element $x\in G$ if the order of multiplication is irrelevant: $gx=xg$. Otherwise said, $g$ conjugates $x$ to itself: $c_g(x)=x$. It natural to ask whether there are any elements which commute with \emph{all others.} There are two very simple cases:
\begin{itemize}
  \item If $G$ is abelian, then every element commutes with everything!
  \item The identity $e$ commutes with everything, regardless of $G$.
\end{itemize}
In general, the set of such elements falls somewhere between these extremes. This subset will turn out to be yet another normal subgroup of $G$.

\begin{defn}{}{}
	The \emph{center} of a group $G$ is the subset of $G$ which commutes with everything in $G$:
	\[
		Z(G):=\{g\in G:\forall h\in G,\ gh=hg\}
	\]
\end{defn}

We will prove that $Z(G)\triangleleft G$ shortly. First we give a few examples; unless $G$ is abelian, the center is typically difficult to compute, so we omit more of the details.


\begin{examples}{}{centers}
	\exstart $Z(G)=G\iff G$ is abelian.
	\begin{enumerate}\setcounter{enumi}{1}
	  \item $Z(S_n)=\{e\}$ if $n\ge 3$. This is Exercise \ref*{sec:perm1}.\ref{exs:centerSn}.
	  \item $Z(D_{2n+1})=\{e\}$ and $Z(D_{2n})=\{e,\rho_{n/2}\}$, where $\rho_{n/2}$ is rotation by $\ang{180}$. Part of this is in Exercise \ref{exs:centerdn}: more generally, this follows from Exercise \ref*{sec:perm1}.\ref{exs:dndefn}.
	  \item $Z\bigl(\rGL_n(\R)\bigr)=\{\lambda I_n:\lambda\in\R^\times\}$. If you've done enough linear algebra to be familiar with eigenvectors, an argument is reasonably straightforward (Exercise \ref{exs:zglnr})
	\end{enumerate}
\end{examples}

\begin{thm}{}{centernormal}
	Let $G$ be any group. Then:
	\begin{enumerate}
	  \item $Z(G)\triangleleft G$ \qquad\qquad 2. \ $\quotient G{Z(G)}\cong \Inn G$
	\end{enumerate}
\end{thm}

\begin{proof}
	Everything follows from defining a suitable homomorphism with the correct kernel!\medbreak
	Define $\phi:G\to\Inn G$ by $\phi(g)=c_g$. This is indeed a homomorphism ($\phi(gh)=\phi(g)\phi(h)$):
	\begin{align*}
		\bigl(\phi(gh)\bigr)(x)&=c_{gh}(x) =(gh)x(gh)^{-1} =g(hx h^{-1})g^{-1}
		=c_g(c_h(x))\\
		&=\bigl(\phi(g)\phi(h)\bigr)(x)	
	\end{align*}
	Now observe that
	\begin{align*}
		g\in\ker\phi&\iff \forall x\in G,\ c_g(x)=gx g^{-1}=x\\
		&\iff g\in Z(G)
	\end{align*}
	By Lemma \ref{lemm:homosub}, $\ker\phi=Z(G)$ is a normal subgroup of $G$.\smallbreak
	To finish, observe that $\phi$ is surjective (definition of $\Inn G$). The 1\st{} isomorphism theorem tells us that
	\[
		\quotient{G}{Z(G)}\cong\image\phi=\Inn G
		\tag*{\qedhere}
	\]
\end{proof}

\goodbreak

\begin{exercises}{}{}
	Key concepts:\quad \emph{Conjugation\quad Conjugacy Classes\quad Cycle Types are Conjugacy Classes ($S_n$)}
	\begin{quote}
		\emph{(Inner) Automorphism\quad Center of a Group}
	\end{quote}
	
	\begin{enumerate}    
	  \item Either find some $\rho\in G$ such that $\rho\sigma\rho^{-1}=\tau$, or explain why no such element exists:
		\begin{enumerate}
	  	\item $\sigma=(1\,2\,3)$, \ $\tau=(1\,3\,2)$ where $G=S_3$.
	  	\item $\sigma=(1\,4\,5\,6)(2\,3)(5\,6)$, \ $\tau=(1\,2\,3\,4)(5\,6)(2\,6)$ where $G=S_6$.
	  	\item $\sigma=(1\,4\,5\,6)(2\,3)(5\,6)$, \ $\tau=(1\,2)(3\,5\,6)$ where $G=S_6$.
		\end{enumerate}
		
		
	  \item Recall Example \ref{ex:conjexample}.1. Find another element $\nu\neq\rho$ for which $\nu\sigma\nu^{-1}=\tau$. For a challenge, determine \emph{how many} distinct such $\nu$ exist!
	  
	  
	  \item Prove Lemma \ref{lemm:conjequiv}. Prove that the relation
	  \[
	  	x\sim y\iff \text{$y$ is conjugate to $x$}
	  \]
	  is an equivalence relation on any group $G$. 
	  
	
		\item\begin{enumerate}
	  	\item Suppose $y$ is conjugate to $x$ in a group $G$. Prove that the orders of $x$ and $y$ are identical.
	  	\item Show that the converse to part (a) is \emph{false} by exhibiting two non-conjugate elements of the same order in some group.
		\end{enumerate}
	
	
		\item Let $H\le G$, fix $a\in G$ and define the \emph{conjugate subgroup} $K=c_a(H)=\{aha^{-1}:h\in H\}$.
	  \begin{enumerate}
	    \item Prove that $K$ is indeed a subgroup of $G$.
	    \item Prove that the function $\psi:H\to K:h\mapsto aha^{-1}$ is an isomorphism of groups.
	    \item If $H\triangleleft G$, what can you say about $c_a(H)$?
	    \item Let $H=\{e,(1\,2)\}\le S_3$ and $a=(1\,2\,3)$. Compute the conjugate subgroup $K=c_a(H)$.
	  \end{enumerate}
	  
	  
		\item In Example \ref*{ex:conjexample}.\ref{ex:conjexample2} we saw that $V=\{e,(1\,2)(3\,4),(1\,3)(2\,4),(1\,4)(2\,3)\}\triangleleft S_4$.
		\begin{enumerate}
	  	\item Show that \emph{normal subgroup} is not transitive by giving an example of a normal subgroup $K\triangleleft V$ which is \emph{not normal} in $S_4$.
	  	
	  	\item How many \emph{other} subgroups does $S_4$ have which are isomorphic to $V$? Why are none of them normal in $S_4$?
	  	
			\item Explain why $\quotient{S_4}V$ is a group of order six. Prove that
			\[
				(1\,2)V(1\,3)V\neq (1\,3)V(1\,2)V
			\]
			Hence conclude that $\quotient{S_4}V\cong S_3$.
			
			\item Why is it obvious that the following six left cosets are distinct.
			\[
				V,\ (1\,2)V,\ (1\,3)V,\ (2\,3)V,\ (1\,2\,3)V,\ (1\,3\,2)V
			\]
			(\emph{Hint: Think about how none of the representatives $a$ of the above cosets move the number 4 and consider $aV=bV\iff b^{-1}a\in V\ldots$})
			
			\item Define an isomorphism $\mu:\quotient{S_4}V\to S_3$ and prove that it is an isomorphism.
	\end{enumerate}
	
	
		\item Prove Lemma \ref{lemm:autoorder}: if $\phi\in\Aut G$ and $x\in G$, then $\phi(x)$ has the same order as $x$.
	  
	  
	  \item Describe all automorphisms of the Klein four-group $V$.\par
	  (\emph{Hint: use the previous question!})
	  
	  
	  \item Recall Exercise \ref*{sec:kerimage}.\ref{exs:totient}. Explain why $\Aut\Z_n\cong\Z_n^\times$.\par
	  (\emph{Hint: consider $\phi_k(x)=kx$ where $\gcd(k,n)=1$ and map $\psi:k\mapsto\phi_k$})
	  
	  
	  \item Let $G$ be a group. Prove directly that $Z(G)\triangleleft G$, \emph{without} using Theorem \ref{thm:centernormal}. That is:
		\begin{enumerate}
	  	\item Prove that $Z(G)$ is closed under the group operation and inverses.
	  	\item Prove that $gZ(G)=Z(G)g$ for all $g\in G$.
		\end{enumerate}
		
% 		
% 		\item\label{exs:centerSn} DONE EARILER IN PERM CHAPTER! Suppose $n\ge 3$ and that $\sigma\in Z(S_n)$.
% 		\begin{enumerate}
% 		  \item By considering $\sigma(1\,2)\sigma^{-1}$, prove that $\{\sigma(1),\sigma(2)\}=\{1,2\}$.
% 		  \item If $\sigma(1)=2$, repeat the calculation with $\sigma(1\,3)\sigma^{-1}$ to obtain a contradiction.
% 		  \item Hence, or otherwise, deduce that $Z(S_n)=\{e\}$.
% 		\end{enumerate}
		
		
		\item\label{exs:zglnr} We identify the center of the general linear group.\par
		The $n\times n$ matrix $A=\scalebox{0.6}{%
		$\begin{pmatrix}
		  0&1&0&\cdots&\cdots&0\\
		  0&0&1&&&0\\
		  0&0&0&&&0\\
		  \vdots&&&\ddots&\ddots&\\
		  0&0&0&&0&1\\
		  0&0&0&\cdots&&0
		\end{pmatrix}$}$ has a single one-dimensional eigenspace: $A\ve_1=\V0$.
		\begin{enumerate}
		  \item Let $B\in Z\bigl(\rGL_n(\R)\bigr)$. Use the fact that $AB=BA$ to prove that $B\ve_1=\lambda\ve_1$ for some $\lambda\neq 0$.
		  \item Let $\vx\in\R^n$ be non-zero and $X$ an invertible matrix for which $X\ve_1=\vx$ (i.e.\ $\vx$ is the 1\st{} column of $X$). Prove that $B\vx=\lambda\vx$.
		  \item Since the observation in part (b) holds for \emph{any} $\vx\in\R^n$, what can we conclude about $B$? What is the group $Z\bigl(\rGL_n(\R)\bigr)$?
		\end{enumerate}
		
		
		\item\label{exs:centerdn}\begin{enumerate}
		  \item Prove that $D_4$ has center $Z(D_4)=\{e,\rho_2\}$, where $\rho_2$ is rotation by $180^\circ$.
		  \item State the cosets of $Z(D_4)$. What is the order of each? Determine whether $\quotient{D_4}{Z(D_4)}$ is isomorphic to $\Z_4$ or to the Klein four-group $V$.
		  \item (Hard) \ Can you find a homomorphism $\phi:D_4\to D_4$ whose kernel is $Z(D_4)$?\par
		  (\emph{Hint: draw a picture and think about doubling angles of rotation and reflection!})
		\end{enumerate}
	
	
	% 	\item Prove that the centers of the following groups are as claimed:
	% 	\begin{enumerate}
	%   	\item 
	%   	\item $Z(A_4)=\{e\}$.
	%   	\item $Z(\mathrm{SL}_2(\R))=\left\{\left(\begin{smallmatrix}
	% \lambda&0\\0&\lambda
	% \end{smallmatrix}\right):\lambda=\pm 1\right\}$.
	% 	\end{enumerate}
	% 	What do the cosets of the center $Z(\mathrm{SL}_2(\R))$ look like in part (c)? That is, if $A\in B\cdot Z(\rSL_2(\R))$, what can you say about $A$?
	
	
	
	% \item The group of \emph{outer automorphisms} of $G$ is the factor group $\operatorname{Out} G=\quotient{\Aut G}{\Inn G}$.
	% \begin{enumerate}
	%   \item Prove that if $G$ is abelian then $\operatorname{Out}G\cong\Z_1$ is trivial.
	%   \item 
	% \end{enumerate}
	
		
	  %\item $Z(\rO_n(\R))=\{\pm I_n\}$.
	
	\end{enumerate}
\end{exercises}





% \item[HWfinal]
% \item Consider the map $\phi:\rO_2(\R)\to\rO_2(\R)$ defined by
% \[\phi(Rot(\theta))=Rot(2\theta),\qquad \phi(Ref(\psi))=Ref(2\psi),\]
% where $Rot(\theta)$ is rotation counter-clockwise through $\theta$, and $Ref(\psi)$ is reflection across the line making angle $\psi/2$ with the $x$-axis.
% \begin{enumerate}
% \item Prove that $\phi$ is a homomorphism and compute its kernel. (\emph{You may use the results from earlier homework sheets and the notes instead of computing $Rot(\theta)Ref(\psi)$, etc., directly.})
% \item Prove that $\phi$ restricts to a homomorphism $\phi:D_{2n}\to D_n$ for any positive integer $n$.
% \item What does the first isomorphism theorem say in the case of part (b)? That is, identify the factor group $\quotient{D_{2n}}{\ker\phi}$.
% \item What about in the case of part (a)? Prove your assertions.
% \end{enumerate}
% 
% 
% 
%   \item Let $X\subseteq G$ be a fixed subset of a group $G$. Define the \emph{normalizer} $N(X)$ of $X$ and \emph{centralizer} $C(X)$ of $X$ as follows
%   \begin{gather*}
% N(X)=\{g\in G:gX=Xg\},\qquad C(X)=\{g\in G:\forall x\in X,\ gx=xg\}.
%   \end{gather*}
% \begin{enumerate}
%   \item Show that $N(X)$ is a subgroup of $G$.
%   \item Write down $C(X)$ and $N(X)$ for the subset $X=\{\mu_1,\mu_2,\mu_3\}\subseteq D_3=\{e,\rho_1,\rho_2,\mu_1,\mu_2,\mu_3\}$.
%   \item Let $n\in N(X)$. Given that $C(X)$ is a subgroup of $N(X)$, show that $n$ has the same left and right cosets with $C(X)$ if and only if
%   \[\forall x\in X,\ \forall c\in C(X),\ ncn^{-1}x=xncn^{-1}.\]
%   \item Using (c), or otherwise, prove that the centralizer of any set $X$ is a normal subgroup of the normalizer of $X$.
%   \item If $X=G$, to what theorem of the course does part (d) reduce? \emph{Part (b) is a hint.}
% \end{enumerate} 


% 
% 
% \item\begin{enumerate}
%   \item $(1,3)\in\Z_2\times\Z_4$ has order $\lcm(2,4)=4$.
%   \item $(4,2,1)\in\Z_6\times\Z_4\times\Z_3$ has order $\lcm(3,2,3)=6$.
%   \item $((123),(15)(234))\in S_3\times S_5$ has order $\lcm(3,6)=6$.
%   \item $(\mu_1,\mu_1)\in D_3\times D_4$ has order $\lcm(2,2)=2$.
% \end{enumerate}
% 
% 
% 
% 
%   
%   \item\begin{enumerate}
%     \item Strictly this requires induction on $x$, the induction step being
%     \[\psi(x+1)=\psi(x)+\psi(1)=kx+k=(k+1)x\pmod n\]
%     \item $\psi_3(37)=3\cdot 37=111=24\cdot 4+15= 15\pmod{24}$ while $\psi_3(1)=3$. Clearly these are unequal. In $\Z_{36}$ we have $1=37$ so, if $\psi_3$ is to be a function we must have $\psi_3(1)=\psi_3(37)$. It follows that $\psi_3$ is not a function (not well-defined).
%     \item $\psi_k$ is well-defined if and only if
%     \begin{align*}
%     \forall x,y\in\Z,\quad\psi_k(x+36y)=\psi_k(x) &\iff \forall x,y,\quad kx+36ky\equiv  kx\pmod{24}\\
%     &\iff \forall y,\quad 36ky\equiv 0\pmod{24}\\
%     &\iff 36k\equiv 0\pmod{24}\\
%     &\iff 3k\equiv 0\pmod{2}\\
%     &\iff k\equiv 0\pmod{2}
%     \end{align*}
%     $\psi_k$ is a homomorphism iff $k\in\{0,2,4,6,8,\ldots,22\}$. There are 12 distinct homomorphisms $\psi:\Z_{36}\to\Z_{24}$.
%     \item None of the homomorphisms are injective: in particular $\psi_k(0)=0=\psi_k(12)$ for all of the homomorphisms. More generally, $\nm{\Z_{36}}>\nm{\Z_{24}}$ so there are no injective functions $\Z_{36}\to\Z_{24}$, let alone homomorphisms.\\
%     None of the homomorphisms are surjective either. In particular it is impossible for $\psi_k(x)$ to ever equal 1.
%     \item We must have $\psi(x)=kx$ for some $k$. This is well-defined iff
%     \begin{align*}
%     \forall x,y\in\Z,\quad\psi(x+12y)=\psi(x) &\iff \forall x,y,\quad kx+12ky\equiv  kx\pmod{24}\\
%     &\iff \forall y,\quad 12ky\equiv 0\pmod{24}\\
%     &\iff 12k\equiv 0\pmod{24}\\
%     &\iff k\equiv 0\pmod{2}
%     \end{align*}
%     Some of these are injective, in particular $\psi_2(x)=2x$. Indeed it is straightforward to see that $\psi_k(x)=kx$ is injective iff $\gcd(k,24)=2$. None are surjective.
%     \item Let $d=\gcd(m,n)$. $\psi_k:\Z_m\to\Z_n$ is a well-defined homomorphism iff
%     \begin{align*}
%     \forall x,y\in\Z,\quad\psi(x+my)=\psi(x) &\iff \forall x,y,\quad kx+mky\equiv  kx\pmod{n}\\
%     &\iff \forall y,\quad kmy\equiv 0\pmod{n}\\
%     &\iff km\equiv 0\pmod{n}\\
%     &\iff k\frac md\equiv 0\pmod{\tfrac nd}\\
%     &\iff k\equiv 0\pmod{\tfrac nd}
%     \end{align*}
%     (\emph{Recall that $\gcd(\frac md,\frac nd)=1\implies \exists \lambda,\mu\in\Z$ such that $\lambda\frac md+\mu\frac nd=1\implies \lambda$ is the multiplicative inverse of $\frac md$ modulo $\frac nd$.})\\
%     There are $d=\gcd(m,n)$ distinct homomorphisms, namely $\psi_0,\psi_{\frac nd},\psi_{\frac{2n}d},\ldots,\psi_{\frac{(d-1)n}d}$. 
%     \item\begin{enumerate}
%       \item Suppose that $\psi:\Z_m\to\Z_n$ is an injective homomorphism. Then $\nm{\image\psi}=m$. However, $\image\psi\le\Z_n$. Since all subgroups of $\Z_n$ have order a divisor of $n$, it follows that $m\mid n$.\\
%       Conversely, suppose that $m\mid n$. Then the homomorphism $\psi_{\frac nm}(x)=\frac nmx$ is easily seen to be injective.
%       \item Suppose that $\psi:\Z_m\to\Z_n$ is a surjective homomorphism. Then $\image\psi=\Z_n$. In particular, there must exist some $x\in\Z_m$ for which $\psi(x)=1$. But then $mx=0$ in $\Z_m$, from which
%       \[0=\psi(0)=\psi(mx)=m\quad\text{in }\Z_n \]
%       Therefore $m\equiv 0\pmod n$, from which we see that $n\mid m$.\\
%       Conversely, if $n\mid m$ then $\gcd(m,n)=n$ which says that \emph{every} $k\in\Z_n$ defines a homomorphism $\psi_k:\Z_m\to\Z_n$. Certainly $\psi_1(x)=x$ is surjective.
% 		\end{enumerate}
% 		\item $\psi:\Z_n\to\Z_n$ is a bijective homomorphism if and only if $\psi(1)$ is a generator of $\Z_n$. The generators of $\Z_n$ are precisely those elements which are relatively prime to $n$: that is, the units.
% 		\item Let $x,y\in\Z_n^\times$. Then $\gcd(x,n)=1=\gcd(y,n)$ which says that
% 		\[\exists \kappa,\lambda,\mu,\nu\in\Z\quad\text{such that}\quad\begin{cases}
% 		\kappa x+\lambda n=1\\
% 		\mu y+\nu n=1
% 		\end{cases}\]
% 		But this says that $\kappa\mu xy=1+n(\lambda\nu-\lambda-\nu)$, from which $\gcd(xy,n)=1$. Thus $\Z_n^\times$ is closed under multiplication.\\
% 		Multiplication is associative, and the identity element $1\in\Z_n$ is certainly a unit.\\
% 		Finally, if $x\in\Z_n^\times$, then $\kappa x+\lambda n=1\implies \kappa x\equiv 1\pmod n$, whence $x^{-1}=\kappa$. 
% 		\item Let $\phi:\Z_n^\times\to\Aut(\Z_n)$ be defined by $\phi(k)=\psi_k$. We have already seen that $\psi_k$ is an isomorphism if and only if $k$ is a unit, so this is a bijective function. Now compute what $\phi(kl)$ does to an element $x\in\Z_n$:
% 		\[\phi(kl)(x)=\psi_{kl}(x)=(kl)x=k(lx)=\left(\phi(k)\circ\phi(l)\right)(x)\]
% 		It follows that $\phi(kl)=\phi(k)\circ\phi(l)$, and so $\phi$ is a homomorphism, and thus an isomorphism.
%   \end{enumerate}

